Follow-up Samples

### 2017-01-12 submission/manuscript_outline_20161024.Rmd
complete intro edit

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].  Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.

This increased incidence of CRC in the younger population is concerncing since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].  This increased risk can also carry with it an increased risk of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].  Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at risk of reccurence.  Concurrently with this there has been a lot of interest in new areas such as the gut bacterial microbiome for insight into potential disease pathology.

There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  There have also been a few studies that have shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved ith the pathogenesis of CRC.  These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC. However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma.

In this study we investigated what happened to the bacterial microbiome before and after surgery for both adenoma and carcinoma individuals.  Our anlaysis includes both alpha and beta diversity analysis along with investigating indvidual operational taxonomic units (OTUs).  We next utilized a Random Forest model that was trained on a completely separate data set and observed how this model as well as specific OTUs within this model performed on the initial and follow up samples.  We then created a new model to classify intial and follow up samples and investigated how the model performs and what specific OUTs within the model change.  Finally, we observed how specific bacteria that have been previously implicated in CRC changed before and after treatment in both the adenoma and carcinoma individuals.

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease, thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology of the disease has not been well worked out [@haggar_colorectal_2009].  Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been an consistent additional increase in the incidence of CRC in the younger population.

This increased incidence of CRC in the younger population is concerning since having either an adenoma or carcinoma increases ones risk for future adenomas or carcinomas [@laiyemo_short-_2013; @matsuda_five-year_2009; @ren_long-term_2016].  This increased risk can also carry with it an increased chance of mortality due to this recurrence [@loberg_long-term_2014; @freeman_natural_2013].  Therefore there has been a great amount of interest in early risk stratification tools [@lee_identification_2016; @richards_evidence-based_2016] that can help identify those they may be at most susceptibility to reccurence.  Concurrently, there has also been a lot of interest in new areas that could have a role in disease pathogenesis, such as the gut bacterial microbiome.

There has been promising work on the bacterial microbiome and it's ability to be able to complement existing screening methods such as Fecal Immunoglobulin Test (FIT) or act alone as a screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014]. There has also been research into how this microbiome could be altered directly on tumor tissue itself [@dejea_microbiota_2014].  A few studies have now even shown how this microbiome [@zackular_manipulation_2016] or specific members within it [@arthur_microbial_2014] could be directly involved with the pathogenesis of CRC.  These studies have helped to provide a tantilzing link between the bacterial microbiome and CRC. However, at this present time there remains limited information on the bacterial microbiome before and after successful surgery for removal of the adenoma or carcinoma and whether it changes at all.

In this study we investigated what happened to the bacterial microbiome before and after surgery for both adenoma and carcinoma individuals.  Our anlaysis includes both alpha and beta diversity analysis along with investigation of individual operational taxonomic units (OTUs).  We also utilized Random Forest models and observed how these models well as specific OTUs within this model performed on before (initial) and after (follow up) surgery samples.  We used these models to look for similar important OTUs to identify the most important ones for not only classifying initial and follow up samples but also lesion or normal.
